Co-moving stars in \textsl{Gaia DR1}

### 2016-11-03 paper/ms.tex
worked on method and sec 4.1

\begin{figure*}[p]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/sep_dvtan.pdf}
  \end{center}
  \caption{%
    Difference in tangential velocity vs. physical separation for candidate
    pairs of stars with separation $< 10$~pc and $|\Delta v_t| < 10$~\kms.
    % Note the points with small velocity difference: these are likely
    % widely-separated binary stars.
    \label{fig:dv-sep}}
\end{figure*}

@@

The abundance of pairs of stars with small velocity difference in
\figname~\ref{fig:dv-sep} suggests that there are a
significant number of co-moving stars in the \tgas\ data at a range
of separations.
Here we develop a method to select high-confidence co-moving
stars that properly incorporates the complex uncertainties associated with the
\gaia\ data. We make the following assumptions in order to construct a
statistical model (a likelihood function with explicit priors on our
parameters):
\begin{itemize}
  \item We assume that the uncertainties in the data---parallax, $\varpi$, and
    two proper motion components, $\propm = (\begin{array}[t]{c c} \mu_\alpha &
    \mu_\delta\end{array})^\mathsf{T}$---are Gaussian with known covariances
    $\mat{C}$.
    \todo{SMOH: mention GAIA paper that says Gaussian is good approximation?}
  \item We assume that the 3-space velocities of a given pair of stars
    $(\vec{v}_i, \vec{v}_j)$ in the \tgas\ sample (relative to the solar system
    barycenter) are either (1) co-moving with velocity drawn from a velocity
    prior $p(\vec{v})$ and  velocity difference drawn from a zero-mean Gaussian
    with velocity dispersion $s$, or (2) individually drawn from a velocity
    prior $p(\vec{v})$.
\end{itemize}

@@

We compute the fully marginalized likelihood (FML) that a given pair of stars
has the same 3-space velocity with a small velocity difference drawn from a
zero-mean Gaussian with variance $s^2$ (hypothesis 1, $\mathcal{L}_1$), and
the FML of the stars having different 3-space velocities (hypothesis 2,
$\mathcal{L}_2$).
We use the FML ratio $\mathcal{L}_1/\mathcal{L}_2$ as a scalar for selecting
candidate co-moving stars, as described below in more detail.
To compute these FMLs, the likelihood functions $L_i, L_j$ are marginalized
over true 3-space velocity and distance for each star in the pair $(i,j)$.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \vec{v} \,
    L_i(\vec{v}, d_i, s^2) \,
    L_j(\vec{v}, d_j, s^2) \,
    p(\vec{v}) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \vec{v}_i \, \dd^3 \vec{v}_j \,
    L_i(\vec{v}_i, d_i, 0) \,
    L_j(\vec{v}_j, d_j, 0) \,
    p(\vec{v}_i) \, p(\vec{v}_j) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j). \label{eq:hyp2}
\end{align}
The marginalization integral for hypothesis 2 can be split into the product of
two simpler integrals $\mathcal{L}_2 = Q_i \, Q_j$ where
\begin{equation}
  Q = \int \, \dd d \, \dd^3 \vec{v} \, L(\vec{v}, d, 0) \, p(\vec{v}) \, p(d\given\varpi)
\end{equation}
If the velocity prior $p(\vec{v})$ is also Gaussian, the integrals over velocity
in both cases can be performed analytically:
\todo{SMOH: is this what we do?}
We use an equal-weight mixture of three isotropic, zero-mean Gaussian
distributions
\begin{equation}
  p(\vec{v}) = \frac{1}{3} \, \sum_{m=1}^3 \, \mathcal{N}(0, \sigma_{v,m}^2)
\end{equation}
with velocity dispersions $(\sigma_{v,1}, \sigma_{v,2}, \sigma_{v,3}) = (8, 32,
128)~\kms$ meant to represent young disk stars, old disk stars, and halo stars.
We derive the relevant expressions in Appendix~\ref{sec:appendix}.
After marginalizing over velocity, the likelihood integrands only depend on
distance; we numerically compute the integrals over the true distances to each
star in a pair, $d_1,d_2$, using Monte Carlo integration with $K$ samples from
the distance posterior pdfs:
\begin{equation}
  \int \, \dd d \, \tilde{L}(d) \, p(d\given\varpi) \approx
    \frac{1}{K} \, \sum_k^K \, \tilde{L}(d_k)
\end{equation}
where $\tilde{L}(d)$ is the velocity-marginalized likelihood function.
Through experimentation, we have found that $K=128$ samples are sufficient for
estimating the above integrals for stars with a wide range in parallax
signal-to-noise.

@@

In this section, we examine the distribution of likelihood ratios
$\ln \mathcal{L}_1 /\mathcal{L}_2$, and come up with a reasonable cut
for this quantities to select co-moving pairs from candidate pairs list.

With these remarks in mind, we discuss the likelihood ratio distribution.

Figure~\ref{fig:likelihoodratios} shows the likelihood ratios for all 2.6 million
candidate pairs.
Because we do not carry over the normalizations of all probabilities involved
in $L_1$ and $L_2$, the absolute values of the likelihood ratio is not
a meaningful quantity.
Given our crude way of coming up with candidate pair list, we expect
a huge number of pairs that are not genuinely co-moving, compared to
the smaller number of pairs that are.
In order to gauge an appropriate cut on the likelihood ratios,
we compute the likelihood ratios for 200,000 random pairs of stars with the same
signal-to-noise ratio cut as candidate pairs.
Additionally, we select from candidate pair list, those with very small
difference in tangential velocities ($\Delta v_t < 0.5$~km/s) in
two minimum signal-to-noise ratio\footnote{lower of the two stars in a pair}
bins (S/N$>32$ and S/N$<32$) that are highly probable
to be genuine co-moving pairs.
As expected, we see that the likelihood ratio $\ln \mathcal{L}_1 /\mathcal{L}_2$
distribution peaks at higher values, favoring hypothesis 1, for highly probable
co-moving pairs than the random pairs.
We also see that the values are even higher for higher minimum signal-to-noise ratio pairs.

Based on the comparison with random pairs, we can come up with a cut
on the likelihood ratios that reduces the cumulative contamination due to random pairs
to a given threshold.
The value of the cut, the resulting number of pairs, and
its fraction of the total number of candidate pairs
depending on the contamination rate are listed in
Table~\ref{tab:candidatecounts}.
Because the likelihood ratio $\ln \mathcal{L}_1 /\mathcal{L}_2$ asymtotically
is chi-square distribution, the power-law fall off makes the number of resulting
pairs sensitive to the cut.

@@

\begin{table}[tbh]
  \caption{Number of pairs}
  \label{tab:candidatecounts}
  \centering

  \begin{tabular}{l|c|c|c|r}
  \hline

  \hline
  \textbf{Contamination} & \textbf{$\ln \mathcal{L}_1 /\mathcal{L}_2$ cut} &
    \textbf{Count} & \textbf{Fraction} & \textbf{$N_\textrm{me}$} \\
  \hline
    0.01    \% & 5.33 & 92458 & 0.462 \% & 5055 \\
    0.001   \% & 6.53 & 13354 & 0.067 \% & 1549 \\
    0.0001  \% & 7.29 &  2832 & 0.014 \% & 582  \\
    0.00001 \% & 7.35 &  2495 & 0.012 \% & 532  \\
  \hline
  \end{tabular}
\end{table}

\begin{figure*}[p]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/likelihoodratios.pdf}
  \end{center}
  \caption{%
    Density histogram of likelihood ratios for candidate pairs and random pairs.
    We also show the distribution for highly probable genuine co-moving pairs with
    $\Delta v_t < 0.5$~km/s in high ($S/N>32$) and low ($S/N<32$) signal-to-noise ratio
    bins. In both $S/N$ bins, the highly probable co-moving pairs are clearly
    separated from random pairs while the distribution of $\ln \mathcal{L}_1 /\mathcal{L}_2$
    for candidate pairs is expected to be a combination of
    a huge number of ``random'' pairs that are not actually co-moving, and
    a smaller number of genuine co-moving pairs with varying $S/N$.
    \label{fig:likelihoodratios}}
\end{figure*}

>>>>

\begin{figure*}[p]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/sep_dvtan.pdf}
  \end{center}
  \caption{%
    Difference in tangential velocity vs. physical separation for candidate
    pairs of stars with separation $< 10$~pc and $|\Delta v_t| < 10$~\kms.
    \label{fig:dv-sep}}
\end{figure*}

@@

The abundance of pairs of stars with small velocity difference in
\figname~\ref{fig:dv-sep} suggests that there are a
significant number of co-moving stars in the \tgas\ data at a range
of separations.
Here we develop a method to select high-confidence co-moving
stars that properly incorporates the complex uncertainties associated with the
\gaia\ data. We make the following assumptions in order to construct a
statistical model (a likelihood function with explicit priors on our
parameters):
\begin{itemize}
  \item We assume that the uncertainties in the data---parallax, $\varpi$, and
    two proper motion components, $\propm = (\begin{array}[t]{c c} \mu_\alpha &
    \mu_\delta\end{array})^\mathsf{T}$---are Gaussian with known covariances
    $\mat{C}$.
    \todo{SMOH: mention GAIA paper that says Gaussian is good approximation?}
    % http://adsabs.harvard.edu/abs/2012A&A...538A..78L
  \item We assume that the 3-space velocities of a given pair of stars
    $(\vec{v}_i, \vec{v}_j)$ in the \tgas\ sample (relative to the solar system
    barycenter) are either (1) co-moving with velocity drawn from a velocity
    prior $p(\vec{v})$ and  velocity difference drawn from a zero-mean Gaussian
    with velocity dispersion $s$, or (2) individually drawn from a velocity
    prior $p(\vec{v})$.
\end{itemize}

@@

We compute the fully marginalized likelihood (FML) that a given pair of stars
has the same 3-space velocity with a small velocity difference drawn from a
zero-mean Gaussian with variance $s^2$ (hypothesis 1, $\mathcal{L}_1$), and
the FML of the stars having different 3-space velocities (hypothesis 2,
$\mathcal{L}_2$).
We use the FML ratio $\mathcal{L}_1/\mathcal{L}_2$ as a scalar for selecting
candidate co-moving stars, as described below in more detail.
To compute these FMLs, the likelihood functions $L_i, L_j$ are marginalized
over true 3-space velocity and distance for each star in the pair $(i,j)$.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \vec{v} \,
    L_i(\vec{v}, d_i, s^2) \,
    L_j(\vec{v}, d_j, s^2) \,
    p(\vec{v}) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \vec{v}_i \, \dd^3 \vec{v}_j \,
    L_i(\vec{v}_i, d_i, 0) \,
    L_j(\vec{v}_j, d_j, 0) \,
    p(\vec{v}_i) \, p(\vec{v}_j) \, p(d_i \given \varpi_i) \, p(d_j \given \varpi_j). \label{eq:hyp2}
\end{align}
The marginalization integral for hypothesis 2 can be split into the product of
two simpler integrals $\mathcal{L}_2 = Q_i \, Q_j$ where
\begin{equation}
  Q = \int \, \dd d \, \dd^3 \vec{v} \, L(\vec{v}, d, 0) \, p(\vec{v}) \, p(d\given\varpi)
\end{equation}
If the velocity prior $p(\vec{v})$ is also Gaussian, the integrals over velocity
in both cases can be performed analytically:
We use a mixture of three isotropic, zero-mean Gaussian distributions
\begin{equation}
  p(\vec{v}) = \sum_{m=1}^3 \, w_m \mathcal{N}(0, \sigma_{v,m}^2)
\end{equation}
with velocity dispersions $(\sigma_{v,1}, \sigma_{v,2}, \sigma_{v,3}) = (15, 30, 50)
~\kms$ and weights $(w_1,w_2,w_3) = (0.3, 0.55, 0.15)$
meant to encompass young thin disk stars to halo stars.
We derive the relevant expressions in Appendix~\ref{sec:appendix}.
After marginalizing over velocity, the likelihood integrands only depend on
distance; we numerically compute the integrals over the true distances to each
star in a pair, $d_1,d_2$, using Monte Carlo integration with $K$ samples from
the distance posterior pdfs:
\begin{equation}
  \int \, \dd d \, \tilde{L}(d) \, p(d\given\varpi) \approx
    \frac{1}{K} \, \sum_k^K \, \tilde{L}(d_k)
\end{equation}
where $\tilde{L}(d)$ is the velocity-marginalized likelihood function.
Through experimentation, we have found that $K=128$ samples are sufficient for
estimating the above integrals for stars with a wide range in parallax
signal-to-noise. \todo{(SMOH: still unsure about this)}

\todo{discuss vorbit}

@@

In this section, we examine the distribution of likelihood ratios
$\ln \mathcal{L}_1 /\mathcal{L}_2$, and come up with a reasonable cut
for this quantity to select co-moving pairs from candidate pairs list.

\begin{figure*}[p]
  \begin{center}
    \includegraphics[width=\textwidth]{figures/likelihoodratios.pdf}
  \end{center}
  \caption{%
    Left: Density histogram of likelihood ratios for candidate pairs and random pairs.
    Along with the distribution for the entire candidate pairs (black thick line),
    we show the distribution for subsets in slices of $\Delta v_t$.
    Right: $\Delta v_t$ vs separation in slices of $\ln \mathcal{L}_1 /\mathcal{L}_2$.
    \label{fig:likelihoodratios}}
\end{figure*}

Figure~\ref{fig:likelihoodratios} shows the likelihood ratios for all
271k candidate pairs examined.
As discussed in \S~\ref{sec:data}, we expect a correlation between likelihood
ratios of pairs, and their distribution in $\Delta v_t$ vs separation plane.
Specifically, as we sweep through from small $\Delta v_t$ to large, we expect
the population of pairs to change from genuinely co-moving to random.
This becomes clear when we look at the distribution of likelihood ratios for
slices of $\Delta v_t$ at 0.5, 5, and 9.5~\kms.
Pairs with $|\Delta v_t-0.5|<0.5$ are most likely actual co-moving pairs, and
their likelihood distribution is narrowly peaked above 6.
The distribution peaks at lower values and gets broader as $\Delta v_t$ increases,
and the number of random pairs increasingly dominate.
On the right panel of Figure~\ref{fig:likelihoodratios}, we show
how the distribution of pairs on $\Delta v_t$ vs separation plane changes
with high to low $\ln \mathcal{L}_1 /\mathcal{L}_2$ ratios.
This is in agreement with our discussion in \S~\ref{sec:data}.
Finally, as a test,
we compute the likelihood ratios for 200,000 random pairs of stars with the same
parallax signal-to-noise ratio cut as candidate pairs ($S/N > 8$).
Shown as gray shaded histogram, this distribution peaks at lower value ($\approx 2$),
and is clearly separated from highly probable co-moving pairs.

Based on these comparisons, we select pairs with
$\ln \mathcal{L}_1 /\mathcal{L}_2 > 6$ as co-moving pair candidates.
Out of 271,232 candidate pairs examined, 13,058 pairs (4.8\%)
satisfy this condition.

@@
